Thesis: draft

### 2021-05-05 draft.tex
latest update

Most of the simple opinion dynamics models make use of a static network in which users are always active and always exchanging opinions/posts with each other. In this kind of model the opinion dynamics might go as follows: each time step a random user is selected after which his/her opinion is updated according to the opinion dynamics model that is implemented. Another possibility is updating all users at the same time during each time step. \\
\newline
It might, however, be a more realistic situation if nodes/users become active/inactive over time. This would be a better representation of real social media platforms, where users are active for some periods of time and then become inactive again. This kind of activation mechanism of the users gives the network a more temporal character. The underlying network structure does not change (in the sense that each user/node has a fixed number of neighbors that does not change over time, so no new friends are made or, conversely, no old friends are being defriended), but by making users active and inactive the exchange of information between users becomes time dependent and depends on which users are active and which are not. Hence, this does affect the network structure in some way, because if a user, for example, does not become active for a long period of time, he/she does not participate in the exchange of information/opinions and the network structure behaves like if this user/node is absent. This may affect the dynamics, since it may seem as if the network has a different average degree and some properties like connectivity, clustering coefficient, etc may behave differently. \\
\newline
In this thesis we follow the same activation mechanism as was proposed in the paper of Perra and Rocha \cite{Perra2019}. Every time step users are made active with some probability $p_i$ (thus, every time step, each node becomes active with a probability $p_i$). In this thesis the activation probability is set equal to $p = 0.1$, which is constant for all the nodes. Perra and Rocha chose, in their paper, activation probabilities that are extracted from a power-law ($F(p) \sim p^{-1.5}$, with $p \in \interval{0.01}{1}$) \cite{Perra2019}. They did this based on the analyses of real social networks that shows that the propensity of users to start an on-line social interaction follows a heterogeneous distribution \cite{Perra2012}\cite{Perra2019}. However, they also showed that using a constant activation probability $p$ over the entire population yields results that are qualitatively similar (at least in their case) \cite{Perra2019}. \\
\newline
For simplicity, this thesis implements the constant activation probability variant. It might however be an interesting extension to switch to the heterogeneous activation probabilities and investigate the possible impacts on the dynamics. \\
\newline
If a user is active, he/she first spreads his current opinion to all his/her neighbors. Then, the active user updates his own opinion according to the opinions on his/her time-line \cite{Perra2019}. This updating goes according to the implemented opinion dynamics model (eg. majority model, probabilistic majority model,...). In every time step the updating of all active nodes is done simultaneously \cite{Perra2019}.

% NOT FOR HERE, but maybe in section over future recommendations, ways forward: talk about need to investigate different activation mechanisms, maybe model where nodes can make new friends and remove old friends (at a time rate comparable to the time rate of the opinion dynamics and also maybe based on the opinions of the friends, eg. if a friend has an opinion far from your, you might be more likely to defriend that one and, vice versa, you might be more likely to become new friends with somebody who has an opinion close to yours)

@@

Ideally, we would test the opinion dynamics model that was used throughout this thesis on network data directly obtained from social network companies. However, most of these companies are not to keen on sharing any form of data. This makes it very hard to find accurate and complete social media network data. Hence, we chose to test the model on other types of social networks, either on-line or off-line. Some possibilities are any form of friendship networks, e-mail networks, collaboration networks, etc. \\
\newline
One of the decisive properties of the real-world networks that were eventually used, was their size. The main, and actually only, reason for this was the computation time. If the networks would be to large, it would take weeks, sometimes even months before there would be any kind of results.\\
\newline
The size of the network models that were used, was around a thousand nodes. For the real-world networks, we also used networks with a size of magnitude of thousand to ten-thousand. This benefits the computation time; it is, however, unknown how the network size affects the results. It is therefore recommended to test the model on larger network models and on larger real-world networks in future researches on this topic. \\
\newline
We used two real-world networks. The data was obtained from \textit{The index of Complex networks} (ICON) \cite{ICON}. The first real-world network is an on-line friendship network among Finnish users of the social music listening and sharing website Last.fm (data from $2006$). Only the largest connected component of this on-line social network is included \cite{ICON}. This data set was originally used in reference \cite{Toivonen2009}, which describes the data set. This on-line social network has the following properties: number of nodes: $N = 8003$; number of edges: $16824$; modularity: $Q = 0.79$; average clustering coefficient: $\left<cc\right> = 0.31$; average degree: $\left<k\right> = 4.2$. The degree distribution is heterogeneous and is best fitted by a log-normal
\begin{equation}
	f(k) = \frac{1}{k \sigma \sqrt{2\pi}} \exp(-\frac{(\ln(k) - \mu)^2}{2 \sigma^2})\ ,
\end{equation}
with $\mu = 1.0078272$ and $\sigma = 0.95173779$. The degree distribution of the Last.fm social network, together with the log-normal fit, is shown in Figure \ref{deg_distr_lastfm}.\\
\newline
The second real-world network that was used is the giant component of the on-line, social network of users of the Pretty-Good-Privacy (PGP) algorithm for secure information interchange (data from $2004$) \cite{ICON}. This network was originally used in reference \cite{Boguna2004}, which gives a detailed description of the network data. The main properties of this network are: number of nodes: $N = 10680$; number of edges: $24340$; modularity: $Q = 0.85$; average clustering coefficient: $\left<cc\right> = 0.27$; average degree: $\left<k\right> = 4.56$. The network has a heterogeneous degree distribution, which is best fitted by two power-laws
\begin{equation}
 f(k) = a k^{-\gamma} \ ,
\end{equation}
with $a = 0.39604951$, $\gamma = 1.53568449$ for $k < 20$ and $a = 5.59021251$, $\gamma = 2.4944057$ for $k \geqslant 20$. The degree distribution for the PGP trust network, together with the power-law fits, is shown in Figure \ref{deg_distr_pgp}.

\begin{figure}[H]
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{Last.fm}
    \includegraphics[width=8cm, height=7cm]{lastfm_degree_distribution_8x7.png}
    \label{deg_distr_lastfm}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{PGP}
    \includegraphics[width=8cm, height=7cm]{PGP_degree_distribution_8x7.png}
    \label{deg_distr_pgp}
  \end{subfigure}
  \captionsetup{format=plain}
  \caption[Degree distribution $P(k)$ of the two real-world networks used in this thesis.]{Degree distribution $P(k)$ of the two real world networks used in this thesis. \textbf{(a)} Last.fm on-line social network with a log-normal fit ($\mu = 1.0078272$, $\sigma = 0.95173779$); \textbf{(b)} PGP on-line trust network with two power-law fits ($a = 0.39604951$, $\gamma = 1.53568449$ for $k < 20$ and $a = 5.59021251$, $\gamma = 2.4944057$ for $k \geqslant 20$).}
\label{ev_op_50_50}
\end{figure}

@@

This chapter is subdivided into four main sections. We first discuss the impact of the network structure on the evolution of the opinions and the formation of echo chambers. Then, we introduce some kind of stubbornness in the system. The impact of how the opinions are initially distributed in a network with community structure (SBM) will be discussed afterwards. Finally, we will run the opinion dynamics model of this thesis on some real-world networks. We will discuss the results and compare them to results obtained from network models with a similar average degree, clustering coefficient and/or community structure.

@@

In this section we will discuss the results related to the network topologies of the different network models that are used in this thesis. We will focus on the effect of a high clustering coefficient and a community structure. Two different opinion distributions are investigated, namely balanced starting conditions and unbalanced starting conditions. For both opinion distributions we investigate the evolution of the prevalence of the opinions with time and the formation of echo chambers. Two filtering algorithms will be compared to each other, namely the REC and the PR filtering algorithm, in order to investigate the possible interplay between those filtering algorithms and the network structures.\\
%\newline
%Before sharing the results, a quick disclaimer: Some of the results that are shared in this section, were already obtained by prof. Nicola Perra and prof. Luis E.C. Rocha in their paper \textit{``Modelling opinion dynamics in the age of algorithmic personalization''} (see reference \cite{Perra2019}). These results include the effect of a high clustering coefficient (Watts-Strogatz model) and a random network model. Similar results (albeit with a different network size and average degree) are included in this section to give a complete overview and to be able to compare a community structure to a high clustering coefficient.

\subsection{Balanced starting conditions}
\label{50-50}
In this configuration $50 \%$ of the nodes start with opinion A ($P_{\text{A}}(0) = 0.5$) and $50 \%$ of the nodes start with opinion B ($P_{\text{B}}(0) = 0.5$). The opinions are randomly distributed over the network. The opinion dynamics model discussed in Chapter \ref{chap4}, Section \ref{opinionmodel} is used. We set the activation probability equal to $p_{\text{act}} = 0.1$ and the actual time-line of each user has a maximum length equal to $S=20$.\\
\newline
Four different network models are implemented, each with a total number of nodes equal to $N = 10^3$ and an average degree around $\left<k\right> \sim 10$. The baseline model is the Erd\H{o}s-R\'{e}nyi model (ER) with an edge probability $p = 0.01$. To investigate the effect of clustering on the dynamics, we use the Watts-Strogatz model (WS). Two different values of the rewire probability $\beta$ are used, each with a slightly different value of the clustering coefficient $\left<cc\right>$: $\beta = 0.01$ with $\left<cc\right> = 0.648 \pm 0.003$ and $\beta = 0.06$ with $\left<cc\right> = 0.557 \pm 0.007$. The impact of community structure on the dynamics is studied with the stochastic block model (SBM) with 10 communities of size 100. Two different modularities are investigated: a high modularity $Q = 0.908 \pm 0.005$ with an edge probability inside the communities $p_{\text{cl}} = 0.1$ and an edge probability between communities $p_{\text{add}} = 0.001$; and a low modularity $Q = 0.220 \pm 0.007$ with $p_{\text{cl}} = 0.03$ and $p_{\text{add}} = 0.008$. Finally, the impact of both clustering and community structure is investigated with the stochastic Watts-Strogatz block model (SBM-WS) with 10 communities of size 100. Each community has the following Watts-Strogatz parameters: mean degree $\left<k\right> =10$ and rewire probability $\beta = 0.01$. Edges are added between communities with a probability $p_{\text{add}} = 0.001$. This is a network with both a high clustering ($\left<cc\right> = 0.553 \pm 0.004$) and a high modularity ($Q = 0.909 \pm 0.004$).  \\
\newline
\textbf{Evolution of the prevalence of the opinions}\\
\newline
We start with the results for the evolution of the opinion fractions in case of an initial $50/50$ opinion distribution. Perra and Rocha showed that, for the case of balanced initial opinions, the prevalence of the opinions do not change over time for neither the random network, nor the Watts-Strogatz network and for none of the different filtering algorithms \cite{Perra2019}. The results for the network model with community structure (SBM, Figure \ref{sbm_50-50}) and the network model with both community structure and high clustering (SBM-WS, Figure \ref{sbm-ws_50-50}) are shown for both the REC and the PR filtering algorithms. A community structure is also not able to break the status quo for both of the filtering algorithms.

\begin{figure}[H]
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{SBM}
    \includegraphics[width=8cm, height=7cm]{fraction_of_opinions_SBM_10x100_01-0001_003-0008_50-50_PR_REC.png}
    \label{sbm_50-50}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{SBM-WS}
    \includegraphics[width=8cm, height=7cm]{fraction_of_opinions_SBM-WS_10x100_10-001-0001_50-50_PR_REC.png}
    \label{sbm-ws_50-50}
  \end{subfigure}
  \captionsetup{format=plain}
  \caption[Evolution of the prevalence of opinion A ($P_{\text{A}}(t)$) over time for an initial $50/50$ opinion distribution.]{Evolution of the prevalence of opinion A ($P_{\text{A}}(t)$) over time for an initial $50/50$ distribution. Two different network models with a community structure are represented, each with average degree $\left<k\right> \sim 10$ and $N = 10^3$: \textbf{(a)} Stochastic block model with 10 communities of size 100, a high modularity case ($p_{\text{cl}} = 0.1,\ p_{\text{add}} = 0.001,\ Q = 0.908 \pm 0.005$) and a low modularity case ($p_{\text{cl}} = 0.03,\ p_{\text{add}} = 0.008,\ Q = 0.220 \pm 0.007$) are depicted; \textbf{(b)} Stochastic Watts-Strogatz block model with 10 communities of size 100, each community has the following WS parameters: $\left<k\right> = 10,\ \beta = 0.01$ and $p_{\text{add}} = 0.001$, $\left<cc\right> = 0.553 \pm 0.004$ and $Q = 0.909 \pm 0.004$. For each model the REC and the PR filtering algorithms are compared. Each curve is the average of $10^2$ independent simulations and data points are shown, together with their standard deviation $\sigma$, for every 50 time steps to improve visualization.}
\label{ev_op_50_50}
\end{figure}
\noindent
\textbf{Echo chambers}\\
\newline
Figure \ref{echo_50_50} shows the distribution of the fraction of friends (nearest neighbors, nn) of node $i$ with the same opinion B at $t=500$, $\left<P_{\text{B}}^{\text{nn}}\right>$. The value shown on the $y$-axis, $F_N(\left<P_{\text{B}}^{\text{nn}}\right>)$, is the result of dividing the average distribution of opinions in each neighborhood at $t=500$ by its corresponding value at $t=0$. The region on the $x$-axis close to zero describes neighborhoods in which none, or very few users, adopt opinion B (these are thus neighborhoods in which most users adopt opinion A). Conversely, the region on the $x$-axis close to one describes neighborhoods in which a majority of the users adopt opinion B. The distribution is obtained by imposing ten equispaced bins, $\Delta x = 0.1$ \cite{Perra2019}. Error bars are omitted for the sake of an improved visualization.

@@

  \caption[Distribution of the fraction of friends (nearest neighbors, nn) of node $i$ with the same opinion B at $t=500$, $\left<P_{\text{B}}^{\text{nn}}\right>$, for an initial $50/50$ opinion distribution.]{Distribution of the fraction of friends (nearest neighbors, nn) of node $i$ with the same opinion B at $t = 500$. The distribution is normalized by dividing for the same quantity computed at $t=0$. \textbf{(a)} REC filtering algorithm; \textbf{(b)} PR filtering algorithm. Four different network models are used, each with $N = 10^3$ and average degree $\left<k\right> \sim 10$: The Erd\H{o}s-R\'{e}nyi model with $p = 0.01$; The stochastic block model with 10 communities of size 100, a high modularity case ($p_{\text{cl}} = 0.1,\ p_{\text{add}} = 0.001,\ Q = 0.908 \pm 0.005$) and a low modularity case ($p_{\text{cl}} = 0.03,\ p_{\text{add}} = 0.008,\ Q = 0.220 \pm 0.007$) are depicted; The stochastic Watts-Strogatz model with 10 communities of size 100, each community has the following WS parameters: $\left<k\right> = 10,\ \beta = 0.01$ and $p_{\text{add}} = 0.001$, $\left<cc\right> = 0.553 \pm 0.004$ and $Q = 0.909 \pm 0.004$; The Watts-Strogatz model with $\left<k\right> = 10$, two slightly different values of the clustering coefficient are used ($\left<cc\right> = 0.557 \pm 0.007$ and $\left<cc\right> = 0.648 \pm 0.003$). The starting conditions are $P_{\text{A}}(0) = 0.5$ and $P_{\text{B}}(0) = 0.5$. Each curve is the average of $10^2$ independent simulations.}

@@

In Figure \ref{echo_50_50} two different filtering algorithms are compared, the REC and the PR filtering algorithms. The results for both filtering algorithms are qualitatively the same. The main difference is that the formation of echo chambers, using the PR filtering algorithm, is about ten times as big compared to the formation of echo chambers using the REC filtering algorithm. \\

@@

The Erd\H{o}s-R\'{e}nyi model (ER) showns no, or a very limited, sign of echo chambers, for both the REC filtering algorithm and the PR filtering algorithm. Networks with a high clustering (WS and SBM-WS), on the other hand, show a clear sign of polarization and formation of echo chambers (visible in the two spikes, one around zero and one around one). Going to slightly higher values of the clustering coefficient ($\left<cc\right> = 0.557 \pm 0.007$ to $\left<cc\right> = 0.648 \pm 0.003$) in the Watts-Strogatz model results in significantly larger echo chambers. This suggests that the formation of echo chambers is really sensitive to changes in the clustering coefficient. The stochastic block model (SBM) with a high modularity, hence, a clear community structure, does show an appearance of polarization (mainly when using the PR filtering algorithm), however, the formation of echo chambers is significantly smaller than compared to networks with a high clustering. The stochastic block model with a low modularity has no, or very limited, echo chambers, similar as for the Erd\H{o}s-R\'{e}nyi   
model. This is to be expected, since a stochastic block model with a low modularity, and hence a very blurred, almost non-existing community structure, strongly resembles an Erd\H{o}s-R\'{e}nyi network. The stochastic Watts-Strogatz block model (SBM-WS) has both a community structure and a high average clustering. The formation of echo chambers when using this model is approximately of the same size as the formation of echo chambers in a Watts-Strogatz model with a similar average clustering. This suggests that it is mainly the high clustering that drives the system to form echo chambers, while the community structure plays a less important role. However, as said before, the stochastic block model with a high modularity also has a (smaller) formation of echo chambers, so having a clear and strong community structure definitely plays some role in phenomena such as polarization and the formation of opinion clusters. \\

@@

For an initial $50/50$ opinion distribution none of the network models, nor the filtering algorithms could break this status quo (Fig. \ref{ev_op_50_50} and related results in reference \cite{Perra2019}). Despite this preserving of the balanced opinion distribution, the interplay of filtering algorithms and network structure can potentially shift the initial randomly distributed opinions to a more clustered opinion distribution and thus lead to polarization and formation of echo chambers.

@@

Figure \ref{ev_op_20_80} shows the results for the evolution of the fraction of nodes with opinion A, which is the minority opinion, over time. For each network model two filtering algorithms are compared, the REC filtering algorithm and the PR filtering algorithm. \\

@@

There are no visible effects of having a clear community structure. Figure \ref{sbm} shows the results for a stochastic block model with a high modularity (clear community structure) and a low modularity (negligible community structure). There is (almost) no difference in the results for a high and a low modularity.

>>>>

Most of the simple opinion dynamics models make use of a static network in which users are always active and always exchanging opinions/posts with each other. In this kind of model the opinion dynamics might go as follows: each time step a random user is selected after which his/her opinion is updated according to the opinion dynamics model that is implemented. Another possibility is updating all users at the same time during each time step. \\
\newline
It might, however, be a more realistic situation if nodes/users become active/inactive over time. This would be a better representation of real social media platforms, where users are active for some periods of time and then become inactive again. This kind of activation mechanism of the users gives the network a more temporal character. The underlying network structure does not change (in the sense that each user/node has a fixed number of neighbors that does not change over time, so no new friends are made or, conversely, no old friends are being defriended), but by making users active and inactive the exchange of information between users becomes time dependent and depends on which users are active and which are not. Hence, this does affect the network structure in some way, because if a user, for example, does not become active for a long period of time, he/she does not participate in the exchange of information/opinions and the network structure behaves like if this user/node is absent. This may affect the dynamics, since it may seem as if the network has a different average degree and some properties like connectivity, clustering coefficient, etc may behave differently. \\
\newline
In this thesis we follow the same activation mechanism as was proposed in the paper of Perra and Rocha \cite{Perra2019}. Every time step users are made active with some probability $p_i$ (thus, every time step, each node becomes active with a probability $p_i$). In this thesis the activation probability is set equal to $p = 0.1$, which is constant for all the nodes. Perra and Rocha chose, in their paper, activation probabilities that are extracted from a power-law ($F(p) \sim p^{-1.5}$, with $p \in \interval{0.01}{1}$) \cite{Perra2019}. They did this based on the analyses of real social networks that shows that the propensity of users to start an on-line social interaction follows a heterogeneous distribution \cite{Perra2012}\cite{Perra2019}. However, they also showed that using a constant activation probability $p$ over the entire population yields results that are qualitatively similar (at least in their case) \cite{Perra2019}. \\
\newline
For simplicity, this thesis implements the constant activation probability variant. It might however be an interesting extension to switch to the heterogeneous activation probabilities and investigate the possible impacts on the dynamics. \\
\newline
If a user is active, he/she first spreads his current opinion to all his/her neighbors. Then, the active user updates his own opinion according to the opinions on his/her time-line \cite{Perra2019}. This updating goes according to the implemented opinion dynamics model (eg. majority model, probabilistic majority model,...). In every time step the updating of all active nodes is done simultaneously \cite{Perra2019}.
\newpage
% NOT FOR HERE, but maybe in section over future recommendations, ways forward: talk about need to investigate different activation mechanisms, maybe model where nodes can make new friends and remove old friends (at a time rate comparable to the time rate of the opinion dynamics and also maybe based on the opinions of the friends, eg. if a friend has an opinion far from your, you might be more likely to defriend that one and, vice versa, you might be more likely to become new friends with somebody who has an opinion close to yours)

@@

Ideally, we would test the opinion dynamics model that was used throughout this thesis on network data directly obtained from social network companies. However, most of these companies are not to keen on sharing any form of data. This makes it very hard to find accurate and complete social media network data. Hence, we chose to test the model on other types of social networks, either on-line or off-line. Some possibilities are any form of friendship networks, e-mail networks, collaboration networks, etc. \\
\newline
One of the decisive properties of the real-world networks that were eventually used, was their size. The main, and actually only, reason for this was the computation time. If the networks would be to large, it would take weeks, sometimes even months before there would be any kind of results.\\
\newline
The size of the network models that were used, was around a thousand nodes. For the real-world networks, we also used networks with a size of magnitude of thousand to ten-thousand. This benefits the computation time; it is, however, unknown how the network size affects the results. It is therefore recommended to test the model on larger network models and on larger real-world networks in future researches on this topic. \\
\newline
We used two real-world networks. The data was obtained from \textit{The index of Complex networks} (ICON) \cite{ICON}. The first real-world network is an on-line friendship network among Finnish users of the social music listening and sharing website Last.fm (data from $2006$). Only the largest connected component of this on-line social network is included \cite{ICON}. This data set was originally used in reference \cite{Toivonen2009}, which describes the data set. This on-line social network has the following properties: number of nodes: $N = 8003$; number of edges: $16824$; modularity: $Q = 0.79$; average clustering coefficient: $\left<cc\right> = 0.31$; average degree: $\left<k\right> = 4.2$. The degree distribution is heterogeneous and is best fitted by a log-normal
\begin{equation}
	f(k) = \frac{1}{k \sigma \sqrt{2\pi}} \exp(-\frac{(\ln(k) - \mu)^2}{2 \sigma^2})\ ,
\end{equation}
with $\mu = 1.0078272$ and $\sigma = 0.95173779$. The degree distribution of the Last.fm social network, together with the log-normal fit, is shown in Fig. \ref{deg_distr_lastfm}.\\
\newline
The second real-world network that was used is the giant component of the on-line, social network of users of the Pretty-Good-Privacy (PGP) algorithm for secure information interchange (data from $2004$) \cite{ICON}. This network was originally used in reference \cite{Boguna2004}, which gives a detailed description of the network data. The main properties of this network are: number of nodes: $N = 10680$; number of edges: $24340$; modularity: $Q = 0.85$; average clustering coefficient: $\left<cc\right> = 0.27$; average degree: $\left<k\right> = 4.56$. The network has a heterogeneous degree distribution, which is best fitted by two power-laws
\begin{equation}
 f(k) = a k^{-\gamma} \ ,
\end{equation}
with $a = 0.39604951$, $\gamma = 1.53568449$ for $k < 20$ and $a = 5.59021251$, $\gamma = 2.4944057$ for $k \geqslant 20$. The degree distribution for the PGP trust network, together with the power-law fits, is shown in Fig. \ref{deg_distr_pgp}.

\begin{figure}[H]
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{Last.fm}
    \includegraphics[width=8cm, height=7cm]{lastfm_degree_distribution_8x7.png}
    \label{deg_distr_lastfm}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{PGP}
    \includegraphics[width=8cm, height=7cm]{PGP_degree_distribution_8x7.png}
    \label{deg_distr_pgp}
  \end{subfigure}
  \captionsetup{format=plain}
  \caption[Degree distribution $P(k)$ of the two real-world networks used in this thesis.]{Degree distribution $P(k)$ of the two real-world networks used in this thesis. \textbf{(a)} Last.fm on-line social network with a log-normal fit ($\mu = 1.0078272$, $\sigma = 0.95173779$); \textbf{(b)} PGP on-line trust network with two power-law fits ($a = 0.39604951$, $\gamma = 1.53568449$ for $k < 20$ and $a = 5.59021251$, $\gamma = 2.4944057$ for $k \geqslant 20$).}
\label{ev_op_50_50}
\end{figure}

@@

This chapter is subdivided into four main sections. We first discuss the impact of the network structure on the evolution of the prevalence of the opinions and the formation of echo chambers. Then, we introduce some kind of stubbornness in the system. The impact of how the opinions are initially distributed in a network with community structure (SBM) will be discussed afterwards. Finally, we will run the opinion dynamics model of this thesis on some real-world networks. We will discuss the results and compare them to results obtained from network models with a similar average degree, clustering coefficient and/or community structure.

@@

In this section we will discuss the results related to the network topologies of the different network models that are used in this thesis. We will focus on the effect of a high clustering coefficient and community structure. Two different opinion distributions are investigated, balanced starting conditions and unbalanced starting conditions. For both starting conditions we investigate the evolution of the prevalence of the opinions with time and the formation of echo chambers. Two filtering algorithms will be compared to each other, the REC and the PR filtering algorithms, in order to investigate the possible interplay between those filtering algorithms and the network structures.\\
%\newline
%Before sharing the results, a quick disclaimer: Some of the results that are shared in this section, were already obtained by prof. Nicola Perra and prof. Luis E.C. Rocha in their paper \textit{``Modelling opinion dynamics in the age of algorithmic personalization''} (see reference \cite{Perra2019}). These results include the effect of a high clustering coefficient (Watts-Strogatz model) and a random network model. Similar results (albeit with a different network size and average degree) are included in this section to give a complete overview and to be able to compare a community structure to a high clustering coefficient.

\subsection{Balanced starting conditions}
\label{50-50}
In this configuration $50 \%$ of the nodes start with opinion A ($P_{\text{A}}(0) = 0.5$) and $50 \%$ of the nodes start with opinion B ($P_{\text{B}}(0) = 0.5$). The opinions are randomly distributed over the network. The opinion dynamics model discussed in Chapter \ref{chap4}, Section \ref{opinionmodel} is used. We set the activation probability equal to $p_{\text{act}} = 0.1$ and the actual time-line of each user has a maximum length equal to $S=20$.\\
\newline
Four different network models are implemented, each with a total number of nodes $N = 10^3$ and an average degree $\left<k\right> \sim 10$. The baseline model is the Erd\H{o}s-R\'{e}nyi model (ER) with an edge probability $p = 0.01$. To investigate the effect of clustering on the dynamics, we use the Watts-Strogatz model (WS). Two different values of the rewire probability $\beta$ are used, each with a slightly different value of the clustering coefficient $\left<cc\right>$: $\beta = 0.01$ with $\left<cc\right> = 0.648 \pm 0.003$ and $\beta = 0.06$ with $\left<cc\right> = 0.557 \pm 0.007$. The impact of community structure on the dynamics is studied with the stochastic block model (SBM) with 10 communities of size 100. Two different modularities are investigated: a high modularity $Q = 0.908 \pm 0.005$ with an edge probability inside the communities $p_{\text{cl}} = 0.1$ and an edge probability between communities $p_{\text{add}} = 0.001$; and a low modularity $Q = 0.220 \pm 0.007$ with $p_{\text{cl}} = 0.03$ and $p_{\text{add}} = 0.008$. Finally, the impact of both clustering and community structure is investigated with the stochastic Watts-Strogatz block model (SBM-WS) with 10 communities of size 100. Each community has the following Watts-Strogatz parameters: mean degree $\left<k\right> =10$ and rewire probability $\beta = 0.01$. Edges are added between communities with a probability $p_{\text{add}} = 0.001$. This is a network with both a high clustering ($\left<cc\right> = 0.553 \pm 0.004$) and a high modularity ($Q = 0.909 \pm 0.004$).  \\
\newline
\textbf{Evolution of the prevalence of the opinions}\\
\newline
We start with the results for the evolution of the prevalence of the opinions in case of an initial $50/50$ opinion distribution. Perra and Rocha showed that, for the case of balanced starting conditions, the prevalence of the opinions do not change over time for neither the random network, nor the Watts-Strogatz network and for none of the filtering algorithms \cite{Perra2019}.

\begin{figure}[H]
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{SBM}
    \includegraphics[width=8cm, height=7cm]{fraction_of_opinions_SBM_10x100_01-0001_003-0008_50-50_PR_REC.png}
    \label{sbm_50-50}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{SBM-WS}
    \includegraphics[width=8cm, height=7cm]{fraction_of_opinions_SBM-WS_10x100_10-001-0001_50-50_PR_REC.png}
    \label{sbm-ws_50-50}
  \end{subfigure}
  \captionsetup{format=plain}
  \caption[Evolution of the prevalence of opinion A ($P_{\text{A}}(t)$) over time for an initial $50/50$ opinion distribution.]{Evolution of the prevalence of opinion A ($P_{\text{A}}(t)$) over time for an initial $50/50$ distribution. Two different network models with community structure are represented, each with average degree $\left<k\right> \sim 10$ and $N = 10^3$: \textbf{(a)} SBM with 10 communities of size 100, a high modularity case ($p_{\text{cl}} = 0.1,\ p_{\text{add}} = 0.001,\ Q = 0.908 \pm 0.005$) and a low modularity case ($p_{\text{cl}} = 0.03,\ p_{\text{add}} = 0.008,\ Q = 0.220 \pm 0.007$) are depicted; \textbf{(b)} SBM-WS with 10 communities of size 100, each community has the following WS parameters: $\left<k\right> = 10,\ \beta = 0.01$ and $p_{\text{add}} = 0.001$, $\left<cc\right> = 0.553 \pm 0.004$ and $Q = 0.909 \pm 0.004$. For each model the REC and the PR filtering algorithms are compared. Each curve is the average of $10^2$ independent simulations and data points are shown for every 50 time steps to improve visualization. Bars represent the standard deviation $\sigma$.}
\label{ev_op_50_50}
\end{figure}
\noindent
The results for the network model with community structure (SBM, Fig. \ref{sbm_50-50}) and the network model with both community structure and high clustering (SBM-WS, Fig. \ref{sbm-ws_50-50}) are shown for both the REC and the PR filtering algorithms. A community structure is also not able to break the status quo for both of the filtering algorithms. The initial prevalence of the opinions is maintained over time.\\
\newline
\textbf{Echo chambers}\\
\newline
Fig. \ref{echo_50_50} shows the distribution of the fraction of friends (nearest neighbors, nn) of node $i$ with the same opinion B at $t=500$, $\left<P_{\text{B}}^{\text{nn}}\right>$. The value shown on the $y$-axis, $F_N(\left<P_{\text{B}}^{\text{nn}}\right>)$, is the result of dividing the average distribution of opinions in each neighborhood at $t=500$ by its corresponding value at $t=0$. The region on the $x$-axis close to zero describes neighborhoods in which none, or very few users, adopt opinion B (these are thus neighborhoods in which most users adopt opinion A). Conversely, the region on the $x$-axis close to one describes neighborhoods in which a majority of the users adopt opinion B. The distribution is obtained by imposing ten equispaced bins, $\Delta x = 0.1$ \cite{Perra2019}.

@@

  \caption[Distribution of the fraction of friends (nearest neighbors, nn) of node $i$ with the same opinion B at $t=500$, $\left<P_{\text{B}}^{\text{nn}}\right>$, for an initial $50/50$ opinion distribution.]{Distribution of the fraction of friends (nearest neighbors, nn) of node $i$ with the same opinion B at $t = 500$. The distribution is normalized by dividing for the same quantity computed at $t=0$. \textbf{(a)} REC filtering algorithm; \textbf{(b)} PR filtering algorithm. Four different network models are used, each with $N = 10^3$ and average degree $\left<k\right> \sim 10$: ER with $p = 0.01$; SBM with 10 communities of size 100, a high modularity case ($p_{\text{cl}} = 0.1,\ p_{\text{add}} = 0.001,\ Q = 0.908 \pm 0.005$) and a low modularity case ($p_{\text{cl}} = 0.03,\ p_{\text{add}} = 0.008,\ Q = 0.220 \pm 0.007$) are depicted; SBM-WS with 10 communities of size 100, each community has the following WS parameters: $\left<k\right> = 10,\ \beta = 0.01$ and $p_{\text{add}} = 0.001$, $\left<cc\right> = 0.553 \pm 0.004$ and $Q = 0.909 \pm 0.004$; WS with $\left<k\right> = 10$, two slightly different values of the clustering coefficient are used ($\left<cc\right> = 0.557 \pm 0.007$ and $\left<cc\right> = 0.648 \pm 0.003$). The starting conditions are $P_{\text{A}}(0) = 0.5$ and $P_{\text{B}}(0) = 0.5$. Each curve is the average of $10^2$ independent simulations. Error bars are omitted for an improved visualization.}

@@

In Fig. \ref{echo_50_50} two different filtering algorithms are compared, the REC and the PR filtering algorithms. The results for both filtering algorithms are qualitatively the same. The main difference is that the formation of echo chambers, using the PR filtering algorithm, is about ten times as big compared to the formation of echo chambers using the REC filtering algorithm. \\

@@

The Erd\H{o}s-R\'{e}nyi model (ER) showns no, or a very limited, sign of echo chambers, for both the REC filtering algorithm and the PR filtering algorithm. Networks with high clustering (WS and SBM-WS), on the other hand, show a clear sign of polarization and formation of echo chambers (visible in the two spikes, one around zero and one around one). Going to slightly higher values of the clustering coefficient ($\left<cc\right> = 0.557 \pm 0.007$ to $\left<cc\right> = 0.648 \pm 0.003$) in the Watts-Strogatz model results in significantly larger echo chambers. This suggests that the formation of echo chambers is really sensitive to changes in the clustering coefficient. The stochastic block model (SBM) with a high modularity, a clear community structure, does show an appearance of polarization (mainly when using the PR filtering algorithm), however, the formation of echo chambers is significantly smaller compared to networks with high clustering. The stochastic block model with a low modularity has no, or very limited, echo chambers, similar as for the Erd\H{o}s-R\'{e}nyi   
model. This is to be expected, since a stochastic block model with a low modularity, and hence a very blurred, almost non-existing community structure, strongly resembles an Erd\H{o}s-R\'{e}nyi network. The stochastic Watts-Strogatz block model (SBM-WS) has both community structure and high average clustering. The formation of echo chambers when using this model is approximately of the same size as the formation of echo chambers in a Watts-Strogatz model with a similar average clustering. This suggests that it is mainly the high clustering that drives the system to form echo chambers, while the community structure plays a less important role. However, as said before, the stochastic block model with a high modularity also has a (smaller) formation of echo chambers, so having a clear and strong community structure definitely plays some role in phenomena such as polarization and the formation of opinion clusters. \\

@@

For balanced starting conditions, none of the network models, nor the filtering algorithms could break the status quo (Fig. \ref{ev_op_50_50} and related results in reference \cite{Perra2019}). Despite this preserving of the balanced opinion distribution, the interplay of filtering algorithms and network structure can potentially shift the initial randomly distributed opinions to a more clustered opinion distribution and thus lead to polarization and formation of echo chambers.

@@

Fig. \ref{ev_op_20_80} shows the results for the evolution of the prevalence of opinion A, which is the minority opinion, over time. For each network model two filtering algorithms are compared, the REC filtering algorithm and the PR filtering algorithm. \\

@@

There are no visible effects of having a clear community structure. Fig. \ref{sbm} shows the results for a stochastic block model with a high modularity (clear community structure) and a low modularity (negligible community structure). There is (almost) no difference in the results for a high and a low modularity.
